---
title: "Compare growing degree days (GDD) from different sources"
author: "Ferrer-Paris, J.R."
date: "February 19, 2021"
output:
  html_document: default
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: console
always_allow_html: true
---

```{r setup, include=FALSE}

## load libraries
require(ncdf4)
require(chron)
require(raster)
require(sf)
require(abind)
require(units)
require(plotly)
require(lattice)
require(forecast)
require(dplyr)
require(reshape2)


## set up data directories and other variables
SCRIPTDIR <- sprintf("/home/%s/proyectos/UNSW/australian-alps-climate-change",system("whoami",intern=T))
MAPS <- sprintf("/srv/scratch/%s/gisdata/aust-alps",system("whoami",intern=T))
RDATA <- sprintf("/srv/scratch/%s/gisdata/aust-alps/Rdata",system("whoami",intern=T))
PERIOD <- "1990-2009"
MODEL <- "CCCMA3.1"
PRM <- "R2"
Sys.setenv('MAPBOX_TOKEN' = readLines("~/.mapbox.token"))

```

# Weather station data


```{r feldmark-shape, include=TRUE}
load(sprintf("%s/BOM-ACORN-data.rda",MAPS))
```



# GDD values from CHELSA v1.2 timeseries

We load the GDD data for the period 1979 to 2013 for Tasmania and the Australian Alps:

```{r load-CHELSA-data, include=TRUE}
tas.GDD <- stack(dir(sprintf("%s/CHELSA/GTS",MAPS),"Tasmania",full.names=T))
alp.GDD <- stack(dir(sprintf("%s/CHELSA/GTS",MAPS),"AustAlps",full.names=T))

q1 <- cellFromXY(raster(alp.GDD,1),wst.xy)
wst.alp <- subset(wst.xy,!is.na(q1))
dts.alp <- cbind(wst.alp@data,values(alp.GDD)[q1[!is.na(q1)],])


dts.alp %>%  gather( raster, GTS0.CHELSA, CHELSA_gts0_1979_V1.2.1.AustAlps:CHELSA_gts0_2013_V1.2.1.AustAlps) %>% mutate(year=gsub("CHELSA_gts0_|_V1.2.1.AustAlps","",raster)) -> CHELSA.GTS0


alp.ALL <- merge(BOM.GDD,CHELSA.GTS0,by=c("stn_num","stn_name","year"))

alp.ALL %>% mutate(yy=as.numeric(as.character(year))) -> alp.ALL

plot(GTS0.CHELSA~GDD0,alp.ALL)

mdl <- lm(GTS0.CHELSA~GDD0+yy,alp.ALL)

summary(mdl)

##boxplot(I(GTS0.CHELSA-GDD0)~yy,data=alp.ALL)
plot(I(GTS0.CHELSA-GDD0)~yy,data=alp.ALL)
abline(h=0,lty=3,col=2,lwd=2)

w <- subset(alp.ALL,stn_name %in% "Cabramurra" & yy %in% 1979:2013)$GDD0
z <- subset(alp.ALL,stn_name %in% "Cabramurra" & yy %in% 1979:2013)$GTS0.CHELSA

hs <- 50 - (2013-1979)
tsGDD <- ts(w, start = c(1979, 1), end = c(2013, 1), frequency = 1)
tsGTS <- ts(z, start = c(1979, 1), end = c(2013, 1), frequency = 1)
fitGDD <- tslm(tsGDD ~ trend)
fitGTS <- tslm(tsGTS ~ trend)
fcstGDD <- forecast(fitGDD, h=hs, level=c(80))


q2 <- cellFromXY(raster(tas.GDD,1),wst.xy)
wst.tas <- subset(wst.xy,!is.na(q2))
dts.tas <- values(tas.GDD)[q2[!is.na(q2)],]

matplot(1979:2013,t(dts.tas))


d1 <- BOM.GDD %>% filter(stn_num %in% wst.alp@data$stn_num) %>% dcast(stn_num~year,value.var="GDD0")


yy <- as.numeric(colnames(d1)[-1])
matplot(yy,t(d1[,-1]),type='l',lty=1,lwd=2,col=1,xlim=c(1970,2020))
matlines(1979:2013,t(dts.alp),lty=2,lwd=2,col=2)


d2 <- BOM.GDD %>% filter(stn_num %in% wst.tas@data$stn_num) %>% mutate(GDD.cor=GDD*365/n) %>%  dcast(stn_num~year,value.var="GDD.cor")
yy <- as.numeric(colnames(d2)[-1])
matplot(yy,t(d2[,-1]),type='l',lty=1,lwd=2,col=1,xlim=c(1970,2020))
matlines(1979:2013,t(dts.tas),lty=2,lwd=2,col=2)

```



We can explore the locations with an interactive map (only in HTML document):

```{r feldmark-plotly-map}

map1 <-
plot_mapbox() %>% add_sf(eco.xy)
plot_mapbox(data.frame(eco.centroids)) %>%
   add_markers(x=~X, y=~Y, color = I("maroon"), size=10,
      name="Feldmark ecosystem")

map1 %>% add_markers(data=data.frame(coordinates(Hm.ll)), x=~V1, y=~V2,
   color = I("green"), size=10, name="H. microcarpa") %>%
   layout(
      mapbox = list(
         zoom = 8,
         center = list(lat = -36.5, lon = 148.5)
      )
   )
```

We query the values of the GDD data at the location of each species and the occurrences of the ecosystem.

```{r CHELSA-data-locs, include=TRUE}

qs <- data.frame()
for (ii in ls(pattern="xy$")) {
   q1 <- get.map.cells(ii,raster(alp.GDD,1))
   if (nrow(q1)>0)
      qs <- rbind(qs,q1)
   rm(q1)
}
qs$spp <- trim(gsub("_|.xy"," ",qs$points))


qs <- data.frame()
for (ii in ls(pattern="xy$")) {
   q1 <- get.map.cells(ii,raster(tas.GDD,1))
   if (nrow(q1)>0)
      qs <- rbind(qs,q1)
   rm(q1)
}
qs$spp <- trim(gsub("_|.xy"," ",qs$points))
dts <- values(tas.GDD)[qs$cell,]
dts.tas <- cbind(qs,dts)

```



```{r CHELSA-data-tas, include=TRUE}
dts.alp$grp <- "other"
dts.alp$grp[dts.alp$spp %in% "eco"] <- "Feldmark"
dts.alp$grp[dts.alp$spp %in% nat.list.alp] <- "diag. spp"
dts.alp$grp[dts.alp$spp %in% c(inv.list.alp,inv.list.tas)] <- "inv. spp"

dts.tas$grp <- "other"
dts.tas$grp[dts.tas$spp %in% "eco"] <- "Feldmark"
dts.tas$grp[dts.tas$spp %in% nat.list.tas] <- "diag. spp"
dts.tas$grp[dts.tas$spp %in% inv.list.tas] <- "inv. spp"

## histogram(~CHELSA_gts0_2010_V1.2.1.AustAlps|grp,dts.alp)
## histogram(~CHELSA_gts0_2010_V1.2.1.Tasmania|grp,dts.tas)

```


```{r CHELSA-data-tas, include=TRUE}


x <- dts.alp$CHELSA_gts0_1980_V1.2.1.AustAlps
ss <- dts.alp$grp %in% "diag. spp"
d1 <-  density(x[ss])
ss <- dts.alp$grp %in% "inv. spp"
d2 <-  density(x[ss])

cf <- sum(d1$y)/sum(d2$y)

 cop.alp <- min(d1$x[d1$y < (d2$y*cf/3)])


 x <- dts.tas$CHELSA_gts0_1980_V1.2.1.Tasmania
 ss <- dts.tas$grp %in% "diag. spp"
 d3 <-  density(x[ss])
 ss <- dts.tas$grp %in% "inv. spp"
 d4 <-  density(x[ss])

 cf <- sum(d3$y)/sum(d4$y)

cop.tas <- min(d1$x[d1$y < (d2$y*cf/3)])

```


```{r CHELSA-data-tas, include=TRUE}


layout(matrix(1:2))
par(mar=c(4,4,4.5,1))
plot(d1,lwd=2,main="NSW",xlab="GDD (>0°C)")
lines(d2$x,d2$y*cf,col=2,lwd=2)
abline(v=cop.alp,lty=3)

plot(d3,lwd=2,main="Tasmania",xlab="GDD (>0°C)")
lines(d4$x,d4$y*cf,col=2,lwd=2)
abline(v=cop.tas,lty=3)


```


```{r CHELSA-data-tas, include=TRUE}

w <- t(dts.alp[dts.alp$grp == "Feldmark",grep("CHELSA",colnames(dts.alp))])

hs <- 50 - (2013-1979)
tsDH <- ts(w, start = c(1979, 1), end = c(2013, 1), frequency = 1)
fit <- tslm(tsDH ~ trend)
fcst <- forecast(fit, h=hs, level=c(80))

plot(NA,xlim=c(1979,2029),ylim=c(0,4000))
boxplot(t(w),at=1979:2013,add=T)
abline(h=cop.alp,lty=3)

#lines(tsDH)
RS <- c()
for (j in 1: length(fcst$forecast)) {
   lines(2014:2029,fcst$forecast[[j]]$mean,col=3)
   lines(2014:2029,fcst$forecast[[j]]$upper,col=2)
   lines(2014:2029,fcst$forecast[[j]]$lower,col=4)
   IV <- mean(tsDH[1:5,j])
   FV <- max(fcst$forecast[[j]]$upper)
   RS <- c(RS,(FV-IV)/(cop-IV))

}
RS[RS<0]
1-ecdf(RS)(c(0.3,.5,.8))

```



# Trend in GDD according to climate models

For one location, we can calculate the trend in annual GDD for the whole time period, and interpolate the expected values for years 2000 and 2050:

```{r interpolate-values, include=TRUE}
y <- unlist(gdd.eco[1,-(1:2)])
x <- as.numeric(colnames(gdd.eco)[-(1:2)])
mdl <- lm(y~x,subset=y>0)
IV <- predict(mdl,data.frame(x=2000))
FV <- predict(mdl,data.frame(x=2050))

plot(x,y,type='n',xlab="Year",ylab="GDD")
rect(2000,-3000,2050,3000,col="palegoldenrod",border="palegoldenrod")
points(x,y,col="maroon",pch=1.2)
abline(mdl,lty=2)
points(2000,IV,pch=19,cex=1.6,type="p")
points(2050,FV,pch=19,cex=1.6,type="p")
text(2000,IV*.9,sprintf("Initial value = %0.2f",IV),adj=c(0,1))
text(2050,FV*1.1,sprintf("Final value = %0.2f",FV),adj=c(1,1))
```

# Calculate relative severity and extent

If we set an *arbitrary* collapse threshold value of $CT=2000$, the relative severity for this location is:

```{r relative-severity, include=TRUE}

CT <- 2000

(FV-IV)/(CT-IV)

```

We can now repeat this for all locations, and calculate the relative severity for all units:

```{r extent-and-severity, include=TRUE}
x <- as.numeric(colnames(gdd.eco)[-(1:2)])
CT <- 2000
eco.RS <- data.frame()
for (k in 1:nrow(gdd.eco)) {
   y <- unlist(gdd.eco[k,-(1:2)])
   mdl <- lm(y~x,subset=y>0)
   IV <- predict(mdl,data.frame(x=2000))
   FV <- predict(mdl,data.frame(x=2050))
   eco.RS <- rbind(eco.RS,data.frame(k,RS=(FV-IV)/(CT-IV)))
}
summary(eco.RS$RS)
```

In this case, relative severity is > 50% for all localities (>80% extent), thus the resulting category would be EN.


# How to determine a meaningful collapse threshold?

We could use reference data to estimate a more appropriate collapse threshold. For example *Hakea microcarpa* was mentioned as a species that could invade Feldmark under rising temperatures. We use occurrence localities from the Atlas of Living Australia as  of an external potentially invasive species)

```{r Hakea-microcarpa, include=TRUE}
arch <- sprintf("%s/Hakea-microcarpa-locs.csv",MAPS)
Hm.xy <- unique(read.table(arch,head=F))
coordinates(Hm.xy) <- 1:2
proj4string(Hm.xy) <-
   '+proj=utm +zone=55 +south +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs'
Hm.ll <- spTransform(Hm.xy,"+init=epsg:4326")
```

The occurrence records are very widespread, in Tasmania the distribution of *Hakea microcarpa* has little overlap with the distribution of the ecosystem, but in the mainland the Feldmark has a very restricted distribution (2 cells from the NARCLiM domain), and it overlaps with records of *H. microcarpa*. We can explore the map interactively:



```{r distribution-all, include=FALSE,echo=FALSE}
# alternative static map:
plot(Hm.ll,col=2,cex=.5)
points(eco.centroids,cex=.5)
plot(Hm.ll,col=2,cex=.5,ylim=c(-37,-36),xlim=c(148,149))
points(unique(GDD[,2:1]),pch=22,cex=8)
points(eco.centroids,cex=.5)
```

We can calculate the expected GDD values for the locations with *Hakea microcarpa*:

```{r GDD-Hakea, include=TRUE}
if (!exists("gdd.col")) {
   xys2 <- coordinates(Hm.ll)
   gdd.col <- data.frame(xys2)

   for (PERIOD in c("1990-2009")) {
      load(sprintf("%s/%s-%s-%s.rda",RDATA,PERIOD,MODEL,PRM))

      for (yy in unique(GDD$year)) {
         ss <- subset(GDD, year %in% yy & n>360 &
            lon>(min(eco.centroids[,1])-1) & lon<(max(eco.centroids[,1])+1) &
               lat>(min(eco.centroids[,2])-1) & lat<(max(eco.centroids[,2])+1))
         if(nrow(ss)>0) {
            dst2 <- pointDistance(ss[,2:1],xys2,lonlat=T,allpairs=T)
            gdd.col[yy] <- apply(dst2,2,function(d) idw(x=ss$GDD,d,p=4.5))
         }
      }
   }
}
valid.col <- apply(!is.na(gdd.col[,-(1:2)]),1,sum)>10
gdd.col <- gdd.col[valid.col,]

head(gdd.col[,1:7])

```

But there is considerable overlap and no clear threshold to separate the ecosystem from the areas with presence of *H. microcarpa*

```{r GDD-Hakea-vs-Feldmark, include=TRUE}
x <- as.numeric(colnames(gdd.col)[-(1:2)])
y <- as.numeric(colnames(gdd.eco)[-(1:2)])

boxplot(gdd.col[,-(1:2)],at=x,col=grey(.8),border=grey(.6))
matpoints(y,t(gdd.eco[,-(1:2)]),pch=1,col="red",cex=1)

```
